---
title: "Math 189 - HW 3"
author: "Maya Lu and Jonathan Li"
date: "5/2/2021"
output: html_document
---

```{r, message = FALSE}
library(ggplot2)
library(lattice)
library(patchwork)
library(dplyr)
library(tidyr)
library(gridExtra)

data <- read.table("hcmv.txt", header = TRUE)
```


## Question 1: Random Scatters

``` {r}
set.seed(2000)
set1 <- sample.int(n = 229345, size = 296, replace = FALSE)
set2 <- sample.int( n = 229345, size = 296, replace = FALSE)
set3 <- sample.int(n = 229345, size = 296, replace = FALSE)
set4 <- sample.int(n = 229345, size = 296, replace = FALSE)
set5 <- sample.int(n = 229345, size = 296, replace = FALSE)
real <- data$location

# dataframe of random scatters
random_scatters <- data.frame(set1, set2, set3, set4, set5, real)

# /////////////////////////////
# VISUALIZE RANDOM SCATTERS
# ////////////////////////////

# ==========
# HISTOGRAM
# =========
binwidth = 4000

random_hist <- function(set, binwidth) {
  ggplot(data = random_scatters, aes(x = set)) +
    geom_histogram(binwidth = binwidth, fill = "steelblue", color = "white") +
    ylab("Frequency") +
    ylim (c(0, 20) )+ 
    xlab("Random") +
    theme_minimal()
}

set1_p =  random_hist(set1, binwidth)
set2_p =  random_hist(set2, binwidth)
set3_p =  random_hist(set3, binwidth)
set4_p =  random_hist(set4, binwidth)
set5_p =  random_hist(set5, binwidth)

real_p = ggplot(data = random_scatters, aes(x = real)) +
    geom_histogram(binwidth = binwidth, fill = "darkgoldenrod3", color = "white") +
    ylab("Frequency") +
    xlab("Palindrome")+
    ylim (c(0, 20) )+ 
    theme_minimal()

grid.arrange(set1_p, set2_p, set3_p, set4_p, set5_p, real_p, 
             top = "Distribution of Locations")

# declutter
rm(set1_p, set2_p, set3_p, set4_p, set5_p, real_p)

```

## Question 2: Location and Spacings

#### Consecutive Hits
Comparing spaces between hits:

```{r}

# calculate spaces between consecutive hits, sort then take the difference
sorted_df <- data.frame(apply(random_scatters, 2, sort))
spaces_df <- data.frame(apply(sorted_df, 2, diff))
spaces_df$position <- 1:295

# define scatterplotting function
spaces_scatter <- function(df, pos, set, ylim, col = "steelblue") {
  ggplot(data = df, aes(x = pos, y = set)) + 
    geom_point(alpha = 0.75, col = col) + 
    xlab("Position") +
    ylab("Number of Spaces") +
    ylim(c(0,ylim)) +
    theme_minimal()
}

# create scatterplots
spaces1 <- spaces_scatter(spaces_df, spaces_df$position, spaces_df$set1, ylim = 6000)
spaces2 <- spaces_scatter(spaces_df, spaces_df$position, spaces_df$set2, ylim = 6000)
spaces3 <- spaces_scatter(spaces_df, spaces_df$position, spaces_df$set3, ylim = 6000)
spaces4 <- spaces_scatter(spaces_df, spaces_df$position, spaces_df$set4, ylim = 6000)
spaces5 <- spaces_scatter(spaces_df, spaces_df$position, spaces_df$set5, ylim = 6000)
spaces6 <- spaces_scatter(spaces_df, spaces_df$position, spaces_df$real, ylim = 6000, col = "darkgoldenrod3")

# plot all four on single window
grid.arrange(spaces1, spaces2, spaces3, spaces4, spaces5, spaces6, ncol = 2,
             top = "Scatterplots of Consecutive Spaces")
rm(spaces1, spaces2, spaces3, spaces4, spaces5, spaces6) # declutter
```


Look at the distribution of hits:

```{r}
# define histogram function
spaces_hist <- function(df, set, binwidth, xlim, ylim, fill = "steelblue"){
  ggplot(data = df, aes(x = set)) + 
    geom_histogram(aes(y = ..density..), binwidth = binwidth, fill = fill, col = "white") + 
    xlim(0, xlim)  +
    ylim(0, ylim) +
    theme_minimal()
} 

# create histogram
hits <- spaces_hist(spaces_df, spaces_df$real, 100, x = 6000, y = 0.0015, fill = "darkgoldenrod3") +
  stat_function(fun=dexp, args = list(rate = 1/mean(spaces_df$real))) +
  geom_density(col = "red") + 
  geom_density(aes(x = set1), col = "dodgerblue") +
  geom_density(aes(x = set2), col = "dodgerblue1") +
  geom_density(aes(x = set3), col = "dodgerblue2") +
  geom_density(aes(x = set4), col = "dodgerblue3") +
  geom_density(aes(x = set5), col = "dodgerblue4") +
  xlab("Consecutive Hits Spacings")

hits
```


Statistically compare to exponential using KS test:

``` {r}
# In a homogenous poisson process, the hits follow a exponential distribution
set1.p <- ks.test(spaces_df$set1, pexp, 1/mean(spaces_df$set1))$p
set2.p <- ks.test(spaces_df$set2, pexp, 1/mean(spaces_df$set2))$p
set3.p <- ks.test(spaces_df$set3, pexp, 1/mean(spaces_df$set3))$p
set4.p <- ks.test(spaces_df$set3, pexp, 1/mean(spaces_df$set4))$p
set5.p <- ks.test(spaces_df$set3, pexp, 1/mean(spaces_df$set5))$p
set6.p <- ks.test(spaces_df$real, pexp, 1/mean(spaces_df$real))$p

data.frame("set1" = set1.p, "set2" = set2.p, 
           "set3" = set3.p,"set4" = set4.p, "set5" = set5.p, "real" = set6.p,
           row.names = c("ks_pval"))

# de-clutter environment
rm(set1.p, set2.p, set3.p, set4.p, set5.p, set6.p)
```


#### Consecutive Pairs
Comparing spaces between consecutive pairs: distance between palindromes that are two hits apart. 
```{r}
# ================================
# COMPARING SPACES BETWEEN PAIRS
# ================================

sum_pairs <- data.frame("set1" = rep(NA, 295), 
                        "set2" = rep(NA, 295), 
                        "set3" = rep(NA, 295), 
                        "set4" = rep(NA, 295),
                        "set5" = rep(NA, 295),
                        "real" = rep(NA, 295))

for (i in 1:295){
  sum_pairs[i,] = sorted_df[i,] + sorted_df[i+1,]
}

spaces_pair <- data.frame(apply(sum_pairs, 2, diff))
spaces_pair$position <- 1:294

# scatterplots
spaces1 <- spaces_scatter(spaces_pair, spaces_pair$position, spaces_pair$set1, ylim = 8000)
spaces2 <- spaces_scatter(spaces_pair, spaces_pair$position, spaces_pair$set2, ylim = 8000)
spaces3 <- spaces_scatter(spaces_pair, spaces_pair$position, spaces_pair$set3, ylim = 8000)
spaces4 <- spaces_scatter(spaces_pair, spaces_pair$position, spaces_pair$real, ylim = 8000, col = "darkgoldenrod3")

grid.arrange(spaces1, spaces2, spaces3, spaces4, ncol = 2, 
             top = "Scatterplot of Consecutive Pair Spacings")
rm(spaces1, spaces2, spaces3, spaces4) # declutter
rm(i)
```

Compare histograms between random hits and palindrome dataset.

```{r, warning = F}
# create histograms
pair <- spaces_hist(spaces_pair, spaces_pair$real, 100, x = 6000, y = 0.0007, fill = "darkgoldenrod3") + 
  stat_function(fun=dgamma, 
                args = list(shape = 2, rate = 1/mean(spaces_df$real))) +
  geom_density(col = "red") + 
  geom_density(aes(x = set1), col = "dodgerblue") +
  geom_density(aes(x = set2), col = "dodgerblue1") +
  geom_density(aes(x = set3), col = "dodgerblue2") +
  geom_density(aes(x = set4), col = "dodgerblue3") +
  geom_density(aes(x = set5), col = "dodgerblue4") +
  xlab("Consecutive Pairs Spacings")

pair
```


Compare to distribution:
```{r, warning = F}
# ks-test
set1.p <- ks.test(spaces_pair$set1, pgamma, 2, 1/mean(spaces_df$set1))$p
set2.p <- ks.test(spaces_pair$set2, pgamma, 2, 1/mean(spaces_df$set2))$p
set3.p <- ks.test(spaces_pair$set3, pgamma, 2, 1/mean(spaces_df$set3))$p
set4.p <- ks.test(spaces_pair$set3, pgamma, 2, 1/mean(spaces_df$set4))$p
set5.p <- ks.test(spaces_pair$set3, pgamma, 2, 1/mean(spaces_df$set5))$p
set6.p <- ks.test(spaces_pair$real, pgamma, 2, 1/mean(spaces_df$real))$p

data.frame("set1" = set1.p, "set2" = set2.p, 
           "set3" = set3.p,"set4" = set4.p, "set5" = set5.p, "real" = set6.p,
           row.names = c("ks_pval"))

rm(set1.p, set2.p, set3.p, set4.p, set5.p, set6.p)
rm(sum_pairs)
rm(spaces_pair)

```

### Consecutive Triplets

```{r}
# ================================
# COMPARING SPACES BETWEEN TRIPLETS
# ================================
sum_triplets <- data.frame("set1" = rep(NA, 294), 
                        "set2" = rep(NA, 294), 
                        "set3" = rep(NA, 294), 
                        "set4" = rep(NA, 294),
                        "set5" = rep(NA, 294),
                        "real" = rep(NA, 294))

for (i in 1:294){
  sum_triplets[i,] = sorted_df[i,] + sorted_df[i+1,] + sorted_df[i+2,]
}

spaces_triplets <- data.frame(apply(sum_triplets, 2, diff))
spaces_triplets$position <- 1:293

# scatter plots
spaces1 <- spaces_scatter(spaces_triplets, spaces_triplets$position, spaces_triplets$set1, ylim = 10000)
spaces2 <- spaces_scatter(spaces_triplets, spaces_triplets$position, spaces_triplets$set2, ylim = 10000)
spaces3 <- spaces_scatter(spaces_triplets, spaces_triplets$position, spaces_triplets$set3, ylim = 10000)
spaces4 <- spaces_scatter(spaces_triplets, spaces_triplets$position, spaces_triplets$real, ylim = 10000,
                          col = "darkgoldenrod3")

grid.arrange(spaces1, spaces2, spaces3, spaces4, ncol = 2)

rm(spaces1, spaces2, spaces3, spaces4)
rm(i)
```


Comparing histograms between random hits and palindrome dataset:

```{r, warning = F}
# create histograms
triplets <- spaces_hist(spaces_triplets, spaces_triplets$real, 100, x = 9000, y = 0.0006, fill = "darkgoldenrod3") + 
  stat_function(fun=dgamma, 
                args = list(shape = 3, rate = 1/mean(spaces_df$real))
  ) + 
  geom_density(col = "red") + 
  geom_density(aes(x = set1), col = "dodgerblue") +
  geom_density(aes(x = set2), col = "dodgerblue1") +
  geom_density(aes(x = set3), col = "dodgerblue2") +
  geom_density(aes(x = set4), col = "dodgerblue3") +
  geom_density(aes(x = set5), col = "dodgerblue4") +
  xlab("Consecutive Triplet Spacings")

triplets
``` 


Compare to distribution using KS statistic
```{r}
# ks-test
set1.p <- ks.test(spaces_triplets$set1, rgamma(10000, 3, 1/mean(spaces_df$set1)))$p
set2.p <- ks.test(spaces_triplets$set2, rgamma(10000, 3, 1/mean(spaces_df$set2)))$p
set3.p <- ks.test(spaces_triplets$set3, rgamma(10000, 3, 1/mean(spaces_df$set3)))$p
set4.p <- ks.test(spaces_triplets$set4, rgamma(10000, 3, 1/mean(spaces_df$set3)))$p
set5.p <- ks.test(spaces_triplets$set5, rgamma(10000, 3, 1/mean(spaces_df$set3)))$p
set6.p <- ks.test(spaces_triplets$real, rgamma(10000, 3, 1/mean(spaces_df$real)))$p

data.frame("set1" = set1.p, "set2" = set2.p, "set3" = set3.p,"set4" = set4.p, "set5" = set5.p, "real" = set6.p,
           row.names = c("ks_pval"))

rm(set1.p, set2.p, set3.p, set4.p, set5.p, set6.p)

```

```{r}
rm(set1.p, set2.p, set3.p, set4.p, set5.p, set6.p)
rm(spaces_triplets, sum_triplets, spaces_df, sorted_df)
rm(sorted_df, spaces_df, spaces_triplets, sum_triplets, triplets, pair, hits )
```

Compare locations
```{r}

# ==========
# Stripplots
# ==========
strip1 <- stripplot(set1, pch=16, cex=0.25, xlab = "Random")
strip2 <- stripplot(set2, pch=16, cex=0.25, xlab = "Random")
strip3 <- stripplot(set3, pch=16, cex=0.25, xlab = "Random")
strip4 <- stripplot(set4, pch=16, cex=0.25, xlab = "Random")
strip5 <- stripplot(set5, pch=16, cex=0.25, xlab = "Random")
strip6 <- stripplot(real, pch=16, cex=0.25, col = "darkgoldenrod3", xlab = "Palindrome")

grid.arrange(strip1, strip2, strip3, strip4, strip5, strip6, ncol = 2, 
             top = "Stripplot of Locations")

#declutter
rm(strip1, strip2, strip3, strip4, strip5, strip6)
```


## Question 3: Counts
```{r}
# GRAPHICAL METHODS

set.seed(2000)
binwidth = 3000
n = 296 # sample size
L = 229345 # total size
M = binwidth # interval length
xlim = (as.integer(L/M) + 1) * M # xlim for hist
n_int = xlim/M

# 1) Histogram of Counts vs Expected Uniform Distribution
expected = n/n_int
se = sqrt(expected / n)
ggplot(data = random_scatters, aes(x = real)) +
    geom_histogram(binwidth = binwidth, fill = "darkgoldenrod3", color = "white") +
    ylab("Frequency") +
    xlab("Palindrome")+
    ylim (c(0, 20) )+ 
    theme_minimal() + 
    geom_hline(yintercept = expected, color = "black") +
    geom_hline(yintercept = expected + (3 * se), color = "darkgray", lty = "dashed") + 
    geom_hline(yintercept = expected - (3 * se), color = "darkgray", lty = "dashed")
```


```{r}

# 2) Plot Residuals
breaks <- seq(0, xlim, by = binwidth)
counts <- hist(data$location, breaks, plot = F)$counts
residuals <- (counts - expected) / sqrt(expected)
to_plot <- data.frame(residuals)

ggplot(to_plot, aes(x = 1:n_int, y = residuals)) + 
  geom_point( col = "steelblue") +
  geom_abline(slope = 0, intercept = 0, color = "black") + 
  geom_abline(slope = 0, intercept = 3, color = "darkred", lty = "dashed") + 
  xlab("interval position") +
  ggtitle("Residual Plot") + 
  theme_minimal()

rm(residuals)
```

```{r}
# 3) Number of Palindromes vs Number of Intervals
pois_random <- rpois(n_int, lambda = mean(counts))
counts_df <- data.frame(counts, pois_random)
to_plot <- gather(counts_df)
lambda = n/n_int


ggplot(to_plot, aes(x= value)) + 
  geom_histogram(aes(fill = key), alpha = 0.5, position = "identity", 
                 binwidth = 1, col = "white") + 
  geom_density(aes( y = ..count.., col = key)) +
  theme_minimal() +
  xlab("Number of Palindromes") +
  ylab("Number of Intervals") +
  scale_fill_discrete(name = "Distribution", labels = c("Observed Palindrome Locations", "Poisson")) +
  scale_color_discrete(name = "Distribution", labels = c("Observed Palindrome Locations", "Poisson"))

plot(
    x = dpois( x = 0:14, lambda),
    type = "line"
  )

lines(density(
  counts_df$counts, 
), col = "blue", add = TRUE)
  
```

Formal Statistical Methods
```{r}
# CHI-SQ TEST: test multiple interval lengths and see if chi-square analysis is consistent with each other

# ---------------
# binwidth = 4000
xlim <- (as.integer(L/4000) + 1) * 4000
n_int = xlim/4000
expected = n/n_int
breaks <- seq(0, xlim, by = 4000)
counts4000 <- hist(data$location, breaks, plot = F)$counts

chisq.stat <- sum((counts - expected) ^ 2 / expected)
d.f = n_int - 1 # uniform distribution so no additional parameters
pval = 1 - pchisq(chisq.stat, df = d.f)
rm(xlim, n_int, expected, breaks, counts4000, d.f)

# ----------------
# binwidth = 3000
xlim <- (as.integer(L/3000) + 1) * 3000
n_int = xlim/3000
expected = n/n_int
breaks <- seq(0, xlim, by = 3000)
counts3000 <- hist(data$location, breaks, plot = F)$counts


chisq.stat <- sum((counts - expected) ^ 2 / expected)
d.f = n_int - 1 # uniform distribution so no additional parameters
pval = 1 - pchisq(chisq.stat, df = d.f)
rm(xlim, n_int, expected, breaks, counts3000, d.f)

# ------------------
# binwidth = 5000
xlim <- (as.integer(L/5000) + 1) * 5000
n_int = xlim/5000
expected = n/n_int
breaks <- seq(0, xlim, by = 5000)
counts5000 <- hist(data$location, breaks, plot = F)$counts


chisq.stat <- sum((counts - expected) ^ 2 / expected)
d.f = n_int - 1 # uniform distribution so no additional parameters
pval = 1 - pchisq(chisq.stat, df = d.f)
rm(xlim, n_int, expected, breaks, counts5000, d.f)


```

Does the distribution of counts follow a poisson distribution:
```{r}
# divide counts into 0, 1, 2, 3, 4, 5, 6, 7, ,8, >=9
counts_df$truncated <- counts_df$counts
counts_df$truncated[counts_df$counts >= 9] <- ">=9"
counts_df$truncated <- as.character(counts_df$truncated)
lvls = c("0", "1", "2", "3", "4", "5", "6", "7", "8", ">=9")
counts_df$truncated <- ordered(counts_df$truncated, levels = lvls)

# get counts of counts
counts <- as.vector(table(counts_df$truncated))

# get expected poisson distribution
xlim <- (as.integer(L/3000) + 1) * 3000
n_int = xlim/3000
lambda = n/n_int
exp_pois <- c(dpois(0:(8), lambda), 1-sum(dpois(0:(8),lambda)))

# chi-sq test
chisq.test(x = counts, p = exp_pois, simulate.p.value = T, B = 10000)
```

### Question 4: The Largest Scatter
```{r}
# figuring out binwidth
plot_bins <- function(dat, palindrome_location, binwidth) {
  return(ggplot(data = dat, aes(x = palindrome_location)) +
           geom_histogram(aes(y = ..density..),
             binwidth = binwidth, fill = "steelblue", color = "white") +
           ggtitle(paste("n = ", binwidth)) +
           ylim(c(0, 0.000020)) +
           theme_minimal() + 
           geom_abline(slope = 0, intercept = (n/(L/binwidth))/n, color = "black")  
         )
}

bin1 <- plot_bins(data, data$location, 2000)
bin2 <- plot_bins(data, data$location, 3000)
bin3 <- plot_bins(data, data$location, 4000)
bin4 <- plot_bins(data, data$location, 5000)
bin5 <- plot_bins(data, data$location, 6000)
bin6 <- plot_bins(data, data$location, 7000)


(bin1 + bin2)/(bin3 + bin4) / (bin5  + bin6) +
    plot_annotation(title = "Distribution of Locations Across Interval Length")

rm(bin1, bin2, bin3, bin4, bin5, bin6)

```

```{r}
# SIMULATION
set.seed(2000)

binwidth = 3000
n = 296 # sample size
L = 229345 # total size
M = binwidth # interval length
xlim = (as.integer(L/M) + 1) * M # xlim for hist
n_int = xlim/M
nreps = 10000

max.counts <- rep(NA, nreps)

for (i in 1:nreps) {
  X = sample.int(n = L, size = n, replace = F)
  X.hist = hist(X, breaks = seq(0, xlim, by = M), plot = F)
  max.counts[i] = max(X.hist$counts)
}

# observed stat
obs_max <- max(hist(data$location,  breaks = seq(0, xlim, by = M), plot = F)$counts)

# plot
to_plot <- data.frame(max.counts)
ggplot(data = to_plot, aes(x = max.counts)) +
  geom_histogram(breaks = seq(6, 14, by = 1) , fill = "steelblue", col = "white") +
  theme_minimal() + 
  geom_vline(xintercept = obs_max, col = "goldenrod3", lty = "dashed") +
  xlab("Max Count") +
  ylab("Frequency") +
  ggtitle("Distribution of Max Counts")

# pvalue
mean(max.counts >= obs_max)
```


### Second Largest Scatter
```{r}
# SIMULATION
set.seed(2000)

binwidth = 3000
n = 296 # sample size
L = 229345 # total size
M = binwidth # interval length
xlim = (as.integer(L/M) + 1) * M # xlim for hist
n_int = xlim/M
nreps = 10000

second.counts <- rep(NA, nreps)

for (i in 1:nreps) {
  X = sample.int(n = L, size = n, replace = F)
  X.hist = hist(X, breaks = seq(0, xlim, by = M), plot = F)
  second.counts[i] = sort(X.hist$counts, decreasing = T)[2]
}

# observed stat
obs_second <- sort((hist(data$location,  breaks = seq(0, xlim, by = M), plot = F)$counts), 
                   decreasing = T)[2]

# plot
to_plot <- data.frame(second.counts)
ggplot(data = to_plot, aes(x = second.counts)) +
  geom_histogram(breaks = seq(6, 12, by = 1) , fill = "steelblue", col = "white") +
  theme_minimal() + 
  geom_vline(xintercept = obs_second, col = "goldenrod3", lty = "dashed") +
  xlab("Second Largest Count") +
  ylab("Frequency") +
  ggtitle("Distribution of Second Largest Scatter")

# pvalue
mean(second.counts >= obs_second)
```